---
title: "p4-explore-and-summarize-data-project"
author: "Courtney Ferguson Lee"
date: "5/9/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Libraries

```{r}
library(ggplot2)
library(tidyr)
library(dplyr)
library(GGally)
library(scales)
library(memisc)
library(readr)
library(gridExtra)
library(RColorBrewer)
library(bitops)
library(RCurl)
#devtools::install_github("dkahle/ggmap")
library(ggmap)
library(maptools)
library(devtools)
#install_github("quantide/mapIT")
#install.packages('choroplethr')
library(choroplethr)
#install.packages('choroplethrMaps')
library(choroplethrMaps)
library(lubridate)
```

## Read Data
Sources of data:

- Uber: https://github.com/fivethirtyeight/uber-tlc-foil-response
- ggmap: D. Kahle and H. Wickham. ggmap: Spatial Visualization with ggplot2. The
R Journal, 5(1), 144-161. URL
  http://journal.r-project.org/archive/2013-1/kahle-wickham.pdf


```{r}
# Read csv's and load taxi lookup data
setwd('/Users/courtneyfergusonlee/p4')
uber <- read.csv('uber-raw-data-janjune-15.csv')
load("uber.RData", .GlobalEnv)
#taxi_lookup <- read.csv('taxi-zone-lookup.csv')

# Get latitude and longitude info for location data
citystate = 'new york city new york'
taxi_lookup <- mutate(taxi_lookup, 
       lat = geocode(paste(Zone, citystate, sep = ', '), output = 'latlon')$lat,
       lon = geocode(paste(Zone, citystate, sep = ', '), output = 'latlon')$lon)
taxi_lookup

# Check for missing latitude or longitude data
taxi_lookup.na <- subset(taxi_lookup, is.na(lat))
taxi_lookup.na

# If there's missing data, need to rename variables and replace values as nec.
names(taxi_lookup.na) <- c("LocationID", 
                           "Borough",    
                           "Zone",       
                           "latitude",
                           "lon")
names(taxi_lookup.na)

# Look up missing values in bulk
taxi_lookup.na <- mutate(taxi_lookup.na, 
       latitude = geocode(paste(Zone, citystate, sep = ', '), 
                          output = 'latlon')$lat)

# Replace missing values individually and CAREFULLY
taxi_lookup[249, 4] <- geocode('West Village, new york city new york', 
                               output = 'latlon')$lat
taxi_lookup[249, ]


# Grab samples to lighten system load
set.seed(420)
uber_sample <- uber[sample(1:length(uber$locationID), 10000), ]
uber_first_stack <- uber[1:10000, ]
uber_bigger_sample <- uber[sample(1:length(uber$locationID), 100000), ]
uber_biggest_sample <- uber[sample(1:length(uber$locationID), 1000000), ]

# Join the sample data with location data
uber_sample <- inner_join(uber_sample, 
                          taxi_lookup, 
                          by = c('locationID' = 'LocationID'))

uber <- inner_join(uber, 
                   taxi_lookup, 
                   by = c('locationID' = 'LocationID'))

```

# (No longer nec) Transform data to get proper densities

This will limit your plots, so planning important. Graph each of the following
1. Bin the data by location
2. Bin the data by location + time:
  - Day
  - Hour
  - Month

```{r}

uber_sample.by_loc <- uber_sample %>%
  group_by(Zone) %>%
  summarise(n = n()) %>%
  inner_join(taxi_copy) %>%
  print

uber_jan.by_loc <- uber_jan %>%
  group_by(Zone) %>%
  summarise(rides = n()) %>%
  inner_join(taxi_copy) %>%
  print

uber_bigger_sample.by_loc <- uber_bigger_sample %>%
  group_by(Zone) %>%
  summarise(n = n()) %>%
  inner_join(taxi_copy) %>%
  print

uber.by_loc <- uber %>%
  group_by(Zone) %>%
  summarise(n = n()) %>%
  inner_join(taxi_copy) %>%
  print

```


## 2d Density Plot of New York Pickups with Scatterplot Layer

Consider getting rid of the density plot.  It doesn't really add much, takes a 
long time to render and seems to break for large sample sizes.  The scatterplot 
seems to be doing a good job of summarizing the data.  It gives actual data 
(real pickup numbers) and can be tweaked if nec. 

```{r}
nyc <- get_map("jackson heights, new york", zoom = 11)

nycMap <- ggmap(nyc,
                    extent = "device", 
                    legend = "none")
# Without density plot (Focus on this.  Extract as much info as possible)
nycMap +
  geom_point(aes(x = lon,
                 y = lat,
                 size = rides,
                 color = rides,
                 alpha = .01),
             #alpha = .01,
             data = subset(uber_jan.by_loc, 
                               Borough != 'Unknown')) +
  scale_color_gradient2(low = 'yellow',
                       high = 'red',
                       mid = 'yellow',
                       guide = guide_colorbar(barheight = 20,
                                              title = 'Pickups')) +
  ggtitle('January Pickups Scatterplot Map')
  scale_size(guide = 'none') +
  guides(alpha = F)


# With density plot (no longer used)
nycMap +
  stat_density2d(aes(x = lon, 
                     y = lat, 
                     fill = ..level..),
                 alpha = .2,
                 bins = 10, 
                 data = subset(uber_sample, 
                               Borough != 'Unknown'),
                 geom = "polygon",
                 show.legend = F) +
  scale_fill_gradient2(#"Pickups",
                       low = "white", 
                       mid = "yellow", 
                       high = "red",
                       limits = c(0, 360),
                       breaks = seq(0, 350, 50)) +
  geom_point(aes(x = lon,
                 y = lat,
                 size = n,
                 color = n,
                 alpha = .01),
             #alpha = .01,
             data = subset(uber_sample.by_loc, 
                               Borough != 'Unknown')) +
  scale_color_gradient2(low = 'yellow',
                       high = 'red',
                       mid = 'yellow',
                       guide = guide_colorbar(barheight = 20,
                                              title = 'Pickups')) +
  scale_size(guide = 'none') +
  guides(alpha = F)

```


Goals:
Done
- Read uber data
- Explore structure
- Get map package (ggmap)
- Grab sample so your system isn't overloaded
- Month by month graphs:
  - Bar graph
  - Freq polygon
  - Box plot
- Make a heatmap for each month
  - Rides by date by month
- Correlation for # rides by:
  - Hour
  - Date
- Heatmap by time:
  - Hour
  - Day
  - Month
- Histogram of times (binned by hour) and number of rides
  - As a line
  - Make multi-level by quartile (.25, .5, .99)
    - Group by day, then do quartile ()
- Histogram of times (binned by hour) and number of rides
  - facet_wrap by day of week
- Analyze:
  - Analyze month cycles:
    - Demand ramps up a hell of a lot throughout the month, esp around the 15th
  - Analyze outliers (April, May and June 100k+)
    Possible Explanations:
      - April was normal
      - May 16: http://www.nyc.gov/html/nypd/html/pr/pr_2015_traffic_2015_05_15.shtml
      - June 27: http://www.nyc.gov/html/nypd/html/pr/pr_2015_traffic_2015_06_26.shtml

Todo
- Fun stuff:
  - Reorganize all tables and graphs
  - Check template
  - Write-up analyses.  Important so far:
    - Festivals are yuuge
    - After the 15th, demand seems to ramp up
    - Pickups pick up over the month
    - Weekday vs weekend patterns

      



## Histograms (Hour, Day, Month)

```{r}

# Create hour, day and month variables
# Note: Don't need to run for samples anymore because 
# it's in the full uber dataset now

uber <- uber %>%
  mutate(hour = hour(Pickup_date),
         day = wday(Pickup_date, label = T, abbr = F),
         month = month(Pickup_date, label = T, abbr = F),
         date = day(Pickup_date))

uber_sample <- uber_sample %>%
  mutate(hour = hour(Pickup_date),
         day = wday(Pickup_date, label = T, abbr = F),
         month = month(Pickup_date, label = T, abbr = F)) %>%
  filter(Zone != 'Unknown')
  
str(uber_sample)
table(uber_sample$Borough)
head(uber_sample$Borough)

# Generate simple graphs by hour, day month
# Use lines for histograms

# Hourly
ggplot(uber_sample,
       aes(x = hour)) +
  geom_freqpoly(binwidth = 1,
                color = I('Blue')) +
  scale_x_continuous(breaks = 0:23) +
  ylab('Number of Pickups')

# Daily
ggplot(uber_sample,
       aes(x = day)) +
  geom_bar(aes(fill = day)) +
  ylab('Number of Pickups') +
  xlab('Day of the Week') +
  scale_fill_brewer(type = 'seq')

# Monthly
# Bar graph of rides per month
ggplot(uber,
       aes(x = month)) +
  geom_bar(aes(fill = month)) +
  scale_y_continuous(labels = comma) +
  scale_fill_brewer(type = 'seq') +
  xlab('Month') +
  ylab('Number of Pickups') +
  guides(guide_legend(title = 'Month'))

# Boxplot of rides per month
ggplot(uber.by_date,
       aes(x = month,
           y = rides)) +
  geom_boxplot(aes(color = month)) +
  scale_color_brewer(type = 'qual',
                     palette = 2) +
  scale_y_continuous(labels = comma) +
  ylab('Number of Pickups') +
  xlab('Month')



## Facets
# Hourly by Day
ggplot(uber_sample,
       aes(x = hour)) +
  geom_freqpoly(binwidth = 1,
                aes(color = day)) +
  scale_x_continuous(breaks = 0:23) +
  scale_color_brewer(type = 'qual', palette = 2) +
  ylab('Number of Pickups') +
  facet_wrap(~ day)

# Hourly by Month
ggplot(uber,
       aes(x = date)) +
  geom_freqpoly(binwidth = 1,
                 aes(color = month)) +
  scale_color_brewer(type = 'qual', palette = 2) +
  scale_y_continuous(label = comma) +
  ylab('Number of Pickups') +
  facet_wrap(~ month)

# Hourly by Borough
ggplot(uber_sample,
       aes(x = hour)) +
  geom_freqpoly(binwidth = 1,
                aes(color = Borough)) +
  scale_x_continuous(breaks = 0:23) +
  scale_color_brewer(type = 'qual', palette = 2) +
  ggtitle('Number of Pickups by Hour and Borough') +
  ylab('Number of Pickups') +
  facet_wrap(~ Borough)

# Daily by Borough
ggplot(uber_jan,
       aes(x = day)) +
  geom_bar(aes(fill = day)) +
  scale_y_log10() +
  ggtitle('Number of Pickups by Day and Borough') +
  ylab('Number of Pickups') +
  xlab('Day of the Week') +
  scale_fill_brewer(type = 'seq') +
  facet_wrap(~ Borough)
ggsave('pickups_by_day_and_borough.png')


# Number of rides per hour scatterplot, color by day, with 3 quartiles
# Make a two separate one with a line for every day (Sun - Sat)

# Three total:
# This plus 1 median line
ggplot(uber_jan.by_hour,
       aes(x = hour,
           y = rides)) +
  geom_point(aes(color = day)) +
  geom_line(stat = 'summary', 
            fun.y = quantile,
            fun.args  =list(probs = .5)) +
  ggtitle('Number of Pickups by Hour and Date') +
  ylab('Number of Pickups') +
  xlab('Time') +
  scale_x_continuous(breaks = 0:23) +
  scale_color_brewer(type = 'div')

# This in one color plus 3 quartiles
ggplot(uber_jan.by_hour,
       aes(x = hour,
           y = rides)) +
  geom_point(aes(color = day)) +
  geom_line(stat = 'summary', fun.y = quantile,
            fun.args  =list(probs = .25), 
            color = 'green') +
  geom_line(stat = 'summary', fun.y = quantile,
            fun.args  =list(probs = .5), 
            color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile,
            fun.args  =list(probs = .95), 
            color = 'purple') +
  #scale_y_log10() +
  ggtitle('Number of Pickups by Hour and Date') +
  ylab('Number of Pickups') +
  xlab('Time') +
  scale_x_continuous(breaks = 0:23) +
  scale_color_brewer(type = 'div')

# This in one color plus 7 days
ggplot(uber_jan.by_hour,
       aes(x = hour,
           y = rides)) +
  geom_point(alpha = .2,
             position = position_jitter()) +
  geom_line(aes(color = day),
            stat = 'summary', 
            fun.y = quantile,
            fun.args  =list(probs = .5)) +
  #scale_y_log10() +
  ggtitle('Number of Pickups by Hour and Date') +
  ylab('Number of Pickups') +
  xlab('Time') +
  scale_x_continuous(breaks = 0:23) +
  scale_color_brewer(type = 'qual', palette = 2)


# Create a 3-graph smoothing progression:

ggplot(uber_jan.by_minute,
       aes(x = hour_plus_minute,
           y = rides)) +
  geom_line()

# By quarter-hour
p1 <- ggplot(uber_jan.by_minute,
       aes(x = round(hour_plus_minute/15, digits = 2)*15,
           y = rides)) +
  geom_line(color = I('green'))

# By half hour
p2<- ggplot(uber_jan.by_minute,
       aes(x = round(hour_plus_minute/30, digits = 2)*30,
           y = rides)) +
  geom_line(alpha = .2)

# By hour
p3 <- ggplot(uber_jan.by_minute,
       aes(x = round(hour_plus_minute/60, digits = 2)*60,
           y = rides)) +
  geom_line(alpha = .2)

grid.arrange(p1, p2, p3, ncol = 1)

# Make a bar graph of Manhattan Zone pickups
ggplot(data = subset(uber_jan,
                     Borough == 'Manhattan'),
       aes(x = Zone)) +
  geom_bar()

subset(uber_sample.by_loc, Borough == 'Manhattan' & n < 50)
```

## Patterns over Time
Look at ride patterns over intervals of time
(Use over Time by: Hour, Day, Month)
Over days (dates), make another frequency polygon

```{r}

# Create month sample (January) and date variable
# Filter out unknown and EWR borough
uber_jan <- uber %>%
  filter(month == 'January' & Borough != 'Unknown' & Borough != 'EWR') %>%
  mutate(date = day(Pickup_date))

# Create histogram of pickups, faceted by date (ncol = 7)
ggplot(uber_jan,
       aes(x = hour)) +
  geom_freqpoly(binwidth = 1,
                aes(color = day)) +
  scale_x_continuous(breaks = 0:23) +
  scale_color_brewer(type = 'qual', palette = 2) +
  ggtitle('January Pickups') +
  ylab('Number of Pickups') +
  facet_wrap(~ date, ncol = 7)

# Look at individual dates:
# June
str(uber.by_date)
ggplot(subset(uber.by_date, month == 'June'),
       aes(x = date,
           y = rides)) +
  geom_col()
  geom_bar() +
  scale_color_brewer(type = 'seq') +
  scale_y_continuous(labels = comma) +
  ylab('Number of Pickups') +
  xlab('Date')

# April
ggplot(subset(uber, 
              month == 'April'),
       aes(x = hour)) +
  geom_freqpoly(binwidth = 1,
                aes(color = day)) +
  scale_x_continuous(breaks = 0:23) +
  scale_color_brewer(type = 'qual', palette = 2) +
  ggtitle('April Pickups') +
  ylab('Number of Pickups') +
  facet_wrap(~ date)

ggplot(uber_jan,
       aes(x = date)) +
  geom_histogram(aes(fill = day),
                 binwidth = 1) +
  scale_fill_brewer(type = 'seq') +
  ggtitle('April Pickups') +
  ylab('Number of Pickups')

# April 25
str(uber)
ggplot(subset(uber, 
              month == 'April'  
              & date == 13),
       aes(x = hour)) +
  geom_freqpoly(binwidth = 1,
                aes(color = day)) +
  scale_x_continuous(breaks = 0:23) +
  scale_color_brewer(type = 'qual', palette = 2) +
  ggtitle('January 13th Pickups') +
  ylab('Number of Pickups')

# May 16
ggplot(subset(uber,
              Month == 'May',
              date == 16),
       aes(x = hour)) +
  geom_freqpoly(binwidth = 1,
                aes(color = day)) +
  scale_x_continuous(breaks = 0:23) +
  scale_color_brewer(type = 'qual', palette = 2) +
  ggtitle('January 13th Pickups') +
  ylab('Number of Pickups')

# June 27
ggplot(subset(uber_jan, date == 13),
       aes(x = hour)) +
  geom_freqpoly(binwidth = 1,
                aes(color = day)) +
  scale_x_continuous(breaks = 0:23) +
  scale_color_brewer(type = 'qual', palette = 2) +
  ggtitle('January 13th Pickups') +
  ylab('Number of Pickups')
# - Tues, Jan 13 (Typical Tuesday)
ggplot(subset(uber_jan, date == 13),
       aes(x = hour)) +
  geom_freqpoly(binwidth = 1,
                aes(color = day)) +
  scale_x_continuous(breaks = 0:23) +
  scale_color_brewer(type = 'qual', palette = 2) +
  ggtitle('January 13th Pickups') +
  ylab('Number of Pickups')
ggsave('jan_13_pickups.png')

# - Fri, Jan 30 (Busy Friday)
ggplot(subset(uber_jan, date == 31),
       aes(x = hour)) +
  geom_freqpoly(binwidth = 1,
                aes(color = day)) +
  scale_x_continuous(breaks = 0:23) +
  scale_color_brewer(type = 'qual', palette = 2) +
  ggtitle('January 31st Pickups') +
  ylab('Number of Pickups')
ggsave('jan_31_pickups.png')

# Bin rides by minute, turning rides into a variable (Do it just)
uber_jan.by_minute <- uber_jan %>%
  mutate(minute = minute(Pickup_date), hour_plus_minute = hour + minute/60) %>%
  group_by(hour_plus_minute) %>%
  summarise(rides = n()) %>%
  ungroup()

# Bin rides by hour, turning rides into a variable.  Need to keep other 
# variables, too
uber_jan.by_hour <- uber_jan %>%
  group_by(date, hour, day) %>%
  summarise(rides = n()) %>%
  ungroup()

uber.by_hour <- uber %>%
  group_by(month, date, hour, day) %>%
  summarise(rides = n()) %>%
  ungroup()

# Testing corellation by hour and date
with(uber.by_hour, cor.test(hour, rides))
with(uber.by_date, cor.test(date, rides))

# Look at number of rides per day in one freq hist
ggplot(uber_jan,
       aes(x = date)) +
  geom_histogram(binwidth = 1,
                aes(fill = day)) +
  scale_x_continuous(breaks = 1:31) +
  scale_fill_brewer(type = 'seq', palette = 1) +
  ggtitle('January Pickups by Date') +
  ylab('Number of Pickups')
ggsave('january_pickups_by_date.png')

# Heatmap by month and date
# Group number of rides by date
uber.by_date <- uber %>%
  group_by(month, date) %>%
  summarise(rides = n())



# Best so far
ggplot(aes(y = date, 
           x = month),
       data = uber.by_date) +
  geom_tile(aes(fill = rides)) +
  scale_fill_gradientn(colours = colorRampPalette(c('blue', 
                                   'white', 
                                   'red'))(100),
                       guide = guide_colorbar(barheight = 20,
                                              title = 'Pickups')) +
  ggtitle('Rides per Date Heatmap') +
  xlab('Month') +
  ylab('Date') +
ggsave('rides_per_date_heatmap.png')

#Trying out scale_fill_gradientn
ggplot(aes(y = date, 
           x = month),
       data = uber.by_date) +
  geom_tile(aes(fill = rides)) +
  scale_fill_gradientn(colors = colorRampPalette(c(                                                   'blue',
                                  'aquamarine4',
                                  "gold",
                                  'firebrick'))(100),
                       guide = guide_colorbar(barheight = 20,
                                              title = 'Pickups')) +
  ylab('Date') +
  xlab('Month') +
  ggtitle('Rides per Date Heatmap')

# Keep going..
ggplot(aes(y = date, 
           x = month),
       data = uber.by_date) +
  geom_tile(aes(fill = rides)) +
  scale_fill_gradientn(colors = colorRampPalette(brewer.pal(11, 'RdYlBu'))(100),
                       guide = guide_colorbar(barheight = 20,
                                              title = 'Pickups')) +
  ylab('Date') +
  xlab('Month') +
  ggtitle('Rides per Date Heatmap')

ggplot(aes(y = date, 
           x = month),
       data = uber.by_date) +
  geom_tile(aes(fill = rides)) +
  scale_fill_gradientn(colors = colorRampPalette(brewer.pal(11, 'Spectral'))(100),
                       guide = guide_colorbar(barheight = 20,
                                              title = 'Pickups')) +
  ylab('Date') +
  xlab('Month') +
  ggtitle('Rides per Date Heatmap')

ggplot(aes(y = date, 
           x = month),
       data = uber.by_date) +
  geom_tile(aes(fill = rides)) +
  scale_colour_manual(values = rev(brewer.pal(11,"Spectral")),
guide = guide_colorbar(barheight = 20,
                                              title = 'Pickups')) +
  ylab('Date') +
  xlab('Month') +
  ggtitle('Rides per Date Heatmap')
```


# Density Map with stat_density2d

```{r}
nyc <- get_map("jackson heights, new york", zoom = 11)

nycMap <- ggmap(nyc,
                    extent = "device", 
                    legend = "none")

nycMap +
  stat_density2d(aes(x = lon, 
                     y = lat, 
                     fill = ..level..),
                 alpha = .2,
                 size = .5, 
                 bins = 50, 
                 data = subset(uber_sample, 
                               Borough != 'Unknown'),
                 geom = "polygon",
                 show.legend = F) +
  scale_fill_gradient2(#"Pickups",
                       #limits = c(0, 200),
                       low = "white", 
                       mid = "yellow", 
                       high = "red",
                       guide = 'none') +
  geom_point(aes(x = lon,
                 y = lat,
                 alpha = .05,
                 color = Borough),
             data = subset(uber_sample, 
                               Borough != 'Unknown'))

```

# Using binned location data

```{r}
uber.by_loc

nyc <- get_map("jackson heights, new york", zoom = 11)
nycMap <- ggmap(nyc,
                    extent = "device")

nycMap +
  geom_point(data = uber.by_loc,
             aes(x = lon,
                 y = lat,
                 color = n,
                 size = n/100,
                 alpha = .1)) +
  coord_trans

```



# Experiments Sections:
- Consider turning map scatterplot into map hexplot

```{r}
nycMap +
  coord_cartesian() +
  stat_binhex(data = subset(uber_sample,
                        Borough != 'Unknown'),
                 aes(x = lon,
                     y = lat,
                     alpha = .05))
?stat_density2d
```


```{r}
data(county.regions)
str(county.regions)
nyc_county_names = c("kings", 
                     "bronx", 
                     "new york", 
                     "queens", 
                     "richmond")
nyc_county_fips = county.regions %>%
  filter(state.name == "new york" & county.name %in% nyc_county_names) %>%
  select(region)
county_choropleth(df_pop_county, 
                  title        = "Population of Counties in New York City",
                  legend       = "Population",
                  num_colors   = 1,
                  county_zoom = nyc_county_fips$region)

```


```{r}
save(taxi_lookup, uber, file = file.path('/Users/courtneyfergusonlee/p4/uber.RData'))
```

