---
title: "Exploratory Data Analysis of Uber Pickup Data"
author: "Courtney Ferguson Lee"
date: "5/9/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r echo = FALSE, message = FALSE, warning = FALSE, packages}
library(ggplot2)
library(tidyr)
library(dplyr)
library(GGally)
library(scales)
library(memisc)
library(readr)
library(gridExtra)
library(RColorBrewer)
library(bitops)
library(RCurl)
#devtools::install_github("dkahle/ggmap")
library(ggmap)
library(maptools)
library(devtools)
#install_github("quantide/mapIT")
library(lubridate)
```



```{r echo=FALSE, Load_the_Data}

# Read csv's and load taxi lookup data
setwd('/Users/courtneyfergusonlee/p4')
uber <- read.csv('uber-raw-data-janjune-15.csv')
taxi_lookup <- read.csv('taxi-zone-lookup.csv')

# May not need this if making this reproducible
#load("uber.RData", .GlobalEnv)

# Get latitude and longitude info for location data (Long Process)
citystate = 'new york city new york'
taxi_lookup <- mutate(taxi_lookup, 
       lat = geocode(paste(Zone, citystate, sep = ', '), output = 'latlon')$lat,
       lon = geocode(paste(Zone, citystate, sep = ', '), output = 'latlon')$lon)

# Check for missing latitude or longitude data
taxi_lookup.lat_na <- subset(taxi_lookup, is.na(lat))
taxi_lookup.lon_na <- subset(taxi_lookup, is.na(lon))
taxi_lookup.lat_na
taxi_lookup.lon_na

# If there's missing data, need to rename variables and replace values as nec.
names(taxi_lookup.lat_na) <- c("LocationID", 
                           "Borough",    
                           "Zone",       
                           "latitude",
                           "longitude")
names(taxi_lookup.lon_na) <- c("LocationID", 
                           "Borough",    
                           "Zone",       
                           "latitude",
                           "longitude")

# Lookup missing values in bulk, check for warnings
taxi_lookup.lat_na <- mutate(taxi_lookup.lat_na, 
       latitude = geocode(paste(Zone, citystate, sep = ', '), 
                          output = 'latlon')$lat)
taxi_lookup.lon_na <- mutate(taxi_lookup.lon_na, 
       longitude = geocode(paste(Zone, citystate, sep = ', '), 
                          output = 'latlon')$lon)

# Replace missing values CAREFULLY (Lookup is a long process)
taxi_lookup[taxi_lookup.lat_na$LocationID, 4] <- taxi_lookup.lat_na$latitude
taxi_lookup[taxi_lookup.lon_na$LocationID, 5] <- taxi_lookup.lon_na$longitude

```



```{r echo=FALSE, Sample_the_Data}

# Grab samples to lighten system load and run tests
set.seed(420)
uber_sample <- uber[sample(1:length(uber$locationID), 100000), ]

# Create Zone, Borough, lat, lon, hour, day, month and date variables
uber_sample <- uber_sample %>%
  inner_join(taxi_lookup, 
             by = c('locationID' = 'LocationID')) %>%
  filter(Zone != 'Unknown' & Zone != 'EWR') %>% # EWR is NJ
  mutate(hour = hour(Pickup_date),
         day = wday(Pickup_date, label = T, abbr = F),
         month = month(Pickup_date, label = T, abbr = F),
         date = day(Pickup_date))

```



```{r echo=FALSE, Format_the_Data}

# Create Zone, Borough, lat, lon, hour, day, month and date variables
uber <- uber %>%
  inner_join(taxi_lookup, 
             by = c('locationID' = 'LocationID')) %>%
  filter(Zone != 'Unknown' & Zone != 'EWR') %>% # EWR is NJ
  mutate(hour = hour(Pickup_date),
         day = wday(Pickup_date, label = T, abbr = F),
         month = month(Pickup_date, label = T, abbr = F),
         date = day(Pickup_date))

# Create month tables for quicker execution
uber_jan <- uber %>%
  filter(month == 'January')
uber_feb <- uber %>%
  filter(month == 'February')
uber_mar <- uber %>%
  filter(month == 'March')
uber_apr <- uber %>%
  filter(month == 'April')
uber_may <- uber %>%
  filter(month == 'May')
uber_jun <- uber %>%
  filter(month == 'June')

# Bin rides by minute, turning rides into a variable (Do it just)
uber_jan.by_minute <- uber_jan %>%
  mutate(minute = minute(Pickup_date), hour_plus_minute = hour + minute/60) %>%
  group_by(hour_plus_minute) %>%
  summarise(rides = n()) %>%
  ungroup()

# Bin rides by hour, turning rides into a variable.  Need to keep other 
# variables, too
uber_jan.by_hour <- uber_jan %>%
  group_by(date, hour, day) %>%
  summarise(rides = n()) %>%
  ungroup()

uber.by_hour <- uber %>%
  group_by(month, date, hour, day) %>%
  summarise(rides = n()) %>%
  ungroup()

# Group number of rides by date
uber.by_date <- uber %>%
  group_by(month, date) %>%
  summarise(rides = n())

# Bin by location for map scatter plot
uber.by_loc <- uber %>%
  group_by(Zone) %>%
  summarise(rides = n()) %>%
  inner_join(taxi_lookup) %>%
  ungroup()


```


# Univariate Plots Section

```{r echo=FALSE, Hourly_Pickups}
ggplot(uber,
       aes(x = hour)) +
  geom_freqpoly(binwidth = 1,
                color = I('Blue')) +
  scale_x_continuous(breaks = 0:23) +
  ggtitle('Frequency Polygon of Pickups per Hour') +
  xlab('Hour') +
  ylab('Number of Pickups')
```




```{r echo=FALSE, Daily_Pickups}
ggplot(uber,
       aes(x = day)) +
  geom_bar(aes(fill = day)) +
  ggtitle('Number of Pickups per Day') +
  ylab('Number of Pickups') +
  xlab('Day of the Week') +
  labs(fill = 'Day') +
  scale_fill_brewer(type = 'seq') +
  ggsave('pickups_per_day_bar.png')
```



```{r echo=FALSE, Monthly Pickups}
# Bar graph of rides per month
ggplot(uber,
       aes(x = month)) +
  geom_bar(aes(fill = month)) +
  scale_y_continuous(labels = comma) +
  scale_fill_brewer(type = 'seq') +
  xlab('Month') +
  ylab('Number of Pickups') +
  guides(guide_legend(title = 'Month'))

# Boxplot of rides per month
ggplot(uber.by_date,
       aes(x = month,
           y = rides)) +
  geom_boxplot(aes(color = month)) +
  scale_color_brewer(type = 'qual',
                     palette = 2) +
  scale_y_continuous(labels = comma) +
  ylab('Number of Pickups') +
  xlab('Month')
```





```{r echo=FALSE, Hourly_Pickups}

```


# Univariate Analysis

## 

### What is the structure of your dataset?

### What is/are the main feature(s) of interest in your dataset?

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

### Did you create any new variables from existing variables in the dataset?

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?



# Multivariate Plots Section

## Map Scatter Plot

```{r echo=FALSE Map_Scatter_Plot}
# Map Scatterplot of New York Pickups
# Create bounding box
lon_left <- min(taxi_lookup$lon)
lon_right <- max(taxi_lookup$lon)
lat_bottom <- min(taxi_lookup$lat)
lat_top <- max(taxi_lookup$lat)

nyc <- get_map(location = c(lon_left, 
                            lat_bottom,
                            lon_right,
                            lat_top), 
               source = 'stamen')

nycMap <- ggmap(nyc,
                    extent = "panel")

nycMap +
  geom_point(aes(x = lon,
                 y = lat,
                 size = rides,
                 color = rides,
                 alpha = .01),
             #alpha = .01,
             data = uber.by_loc) +
  scale_color_gradient2(low = 'yellow',
                       high = 'red',
                       mid = 'yellow',
                       labels = comma,
                       guide = guide_colorbar(barheight = 20,
                                              title = 'Pickups')) +
  ggtitle('Uber Pickups Scatterplot Map') +
  scale_size(guide = 'none') +
  guides(alpha = F) +
  ggsave('scatterplot_map.png')

```


Goals:
Done
- Read uber data
- Explore structure
- Get map package (ggmap)
- Grab sample so your system isn't overloaded
- Month by month graphs:
  - Bar graph
  - Freq polygon
  - Box plot
- Make a heatmap for each month
  - Rides by date by month
- Correlation for # rides by:
  - Hour
  - Date
- Heatmap by time:
  - Hour
  - Day
  - Month
- Histogram of times (binned by hour) and number of rides
  - As a line
  - Make multi-level by quartile (.25, .5, .99)
    - Group by day, then do quartile ()
- Histogram of times (binned by hour) and number of rides
  - facet_wrap by day of week
- Analyze:
  - Analyze month cycles:
    - Demand ramps up a hell of a lot throughout the month, esp around the 15th
  - Analyze outliers (April, May and June 100k+)
    Possible Explanations:
      - April was normal
      - May 16: http://www.nyc.gov/html/nypd/html/pr/pr_2015_traffic_2015_05_15.shtml
      - June 27: http://www.nyc.gov/html/nypd/html/pr/pr_2015_traffic_2015_06_26.shtml

Todo
- Fun stuff:
  - Reorganize all tables and graphs
  - Check template
  - Write-up analyses.  Important so far:
    - Festivals are yuuge
    - After the 15th, demand seems to ramp up
    - Pickups pick up over the month
    - Weekday vs weekend patterns

      



## Simple Graphs

```{r}




## Facets
# Hourly by Day
ggplot(uber_sample,
       aes(x = hour)) +
  geom_freqpoly(binwidth = 1,
                aes(color = day)) +
  scale_x_continuous(breaks = 0:23) +
  scale_color_brewer(type = 'qual', palette = 2) +
  ylab('Number of Pickups') +
  facet_wrap(~ day)

# Hourly by Month
ggplot(uber,
       aes(x = date)) +
  geom_freqpoly(binwidth = 1,
                 aes(color = month)) +
  scale_color_brewer(type = 'qual', palette = 2) +
  scale_y_continuous(label = comma) +
  ylab('Number of Pickups') +
  facet_wrap(~ month)

# Hourly by Borough
ggplot(uber_sample,
       aes(x = hour)) +
  geom_freqpoly(binwidth = 1,
                aes(color = Borough)) +
  scale_x_continuous(breaks = 0:23) +
  scale_color_brewer(type = 'qual', palette = 2) +
  ggtitle('Number of Pickups by Hour and Borough') +
  ylab('Number of Pickups') +
  facet_wrap(~ Borough)

# Daily by Borough
ggplot(uber_jan,
       aes(x = day)) +
  geom_bar(aes(fill = day)) +
  scale_y_log10() +
  ggtitle('Number of Pickups by Day and Borough') +
  ylab('Number of Pickups') +
  xlab('Day of the Week') +
  scale_fill_brewer(type = 'seq') +
  facet_wrap(~ Borough)
ggsave('pickups_by_day_and_borough.png')


# Number of rides per hour scatterplot, color by day, with 3 quartiles
# Make a two separate one with a line for every day (Sun - Sat)

# Three total:
# This plus 1 median line
ggplot(uber_jan.by_hour,
       aes(x = hour,
           y = rides)) +
  geom_point(aes(color = day)) +
  geom_line(stat = 'summary', 
            fun.y = quantile,
            fun.args  =list(probs = .5)) +
  ggtitle('Number of Pickups by Hour and Date') +
  ylab('Number of Pickups') +
  xlab('Time') +
  scale_x_continuous(breaks = 0:23) +
  scale_color_brewer(type = 'div')

# This in one color plus 3 quartiles
ggplot(uber_jan.by_hour,
       aes(x = hour,
           y = rides)) +
  geom_point(aes(color = day)) +
  geom_line(stat = 'summary', fun.y = quantile,
            fun.args  =list(probs = .25), 
            color = 'green') +
  geom_line(stat = 'summary', fun.y = quantile,
            fun.args  =list(probs = .5), 
            color = 'blue') +
  geom_line(stat = 'summary', fun.y = quantile,
            fun.args  =list(probs = .95), 
            color = 'purple') +
  #scale_y_log10() +
  ggtitle('Number of Pickups by Hour and Date') +
  ylab('Number of Pickups') +
  xlab('Time') +
  scale_x_continuous(breaks = 0:23) +
  scale_color_brewer(type = 'div')

# This in one color plus 7 days
ggplot(uber_jan.by_hour,
       aes(x = hour,
           y = rides)) +
  geom_point(alpha = .2,
             position = position_jitter()) +
  geom_line(aes(color = day),
            stat = 'summary', 
            fun.y = quantile,
            fun.args  =list(probs = .5)) +
  #scale_y_log10() +
  ggtitle('Number of Pickups by Hour and Date') +
  ylab('Number of Pickups') +
  xlab('Time') +
  scale_x_continuous(breaks = 0:23) +
  scale_color_brewer(type = 'qual', palette = 2)


# Create a 3-graph smoothing progression:

ggplot(uber_jan.by_minute,
       aes(x = hour_plus_minute,
           y = rides)) +
  geom_line()

# By quarter-hour
p1 <- ggplot(uber_jan.by_minute,
       aes(x = round(hour_plus_minute/15, digits = 2)*15,
           y = rides)) +
  geom_line(color = I('green'))

# By half hour
p2<- ggplot(uber_jan.by_minute,
       aes(x = round(hour_plus_minute/30, digits = 2)*30,
           y = rides)) +
  geom_line(alpha = .2)

# By hour
p3 <- ggplot(uber_jan.by_minute,
       aes(x = round(hour_plus_minute/60, digits = 2)*60,
           y = rides)) +
  geom_line(alpha = .2)

grid.arrange(p1, p2, p3, ncol = 1)

# Make a histogram! of Manhattan Zone pickups
ggplot(data = subset(uber_jan,
                     Borough == 'Manhattan'),
       aes(x = Zone)) +
  geom_histogram()


```

## Patterns over Time
Look at ride patterns over intervals of time
(Use over Time by: Hour, Day, Month)
Over days (dates), make another frequency polygon

```{r}



# Create histogram of pickups, faceted by date (ncol = 7)
ggplot(uber_jan,
       aes(x = hour)) +
  geom_freqpoly(binwidth = 1,
                aes(color = day)) +
  scale_x_continuous(breaks = 0:23) +
  scale_color_brewer(type = 'qual', palette = 2) +
  ggtitle('January Pickups') +
  ylab('Number of Pickups') +
  facet_wrap(~ date, ncol = 7)

# Look at individual dates:
# June
ggplot(subset(uber.by_date, month == 'June'),
       aes(x = date,
           y = rides)) +
  geom_col()
  geom_bar() +
  scale_color_brewer(type = 'seq') +
  scale_y_continuous(labels = comma) +
  ylab('Number of Pickups') +
  xlab('Date')

# April
ggplot(subset(uber, 
              month == 'April'),
       aes(x = hour)) +
  geom_freqpoly(binwidth = 1,
                aes(color = day)) +
  scale_x_continuous(breaks = 0:23) +
  scale_color_brewer(type = 'qual', palette = 2) +
  ggtitle('April Pickups') +
  ylab('Number of Pickups') +
  facet_wrap(~ date)

ggplot(uber_jan,
       aes(x = date)) +
  geom_histogram(aes(fill = day),
                 binwidth = 1) +
  scale_fill_brewer(type = 'seq') +
  ggtitle('April Pickups') +
  ylab('Number of Pickups')

# April 25
ggplot(subset(uber, 
              month == 'April'  
              & date == 13),
       aes(x = hour)) +
  geom_freqpoly(binwidth = 1,
                aes(color = day)) +
  scale_x_continuous(breaks = 0:23) +
  scale_color_brewer(type = 'qual', palette = 2) +
  ggtitle('January 13th Pickups') +
  ylab('Number of Pickups')

# May 16
ggplot(subset(uber,
              Month == 'May',
              date == 16),
       aes(x = hour)) +
  geom_freqpoly(binwidth = 1,
                aes(color = day)) +
  scale_x_continuous(breaks = 0:23) +
  scale_color_brewer(type = 'qual', palette = 2) +
  ggtitle('January 13th Pickups') +
  ylab('Number of Pickups')

# June 27
ggplot(subset(uber_jan, date == 13),
       aes(x = hour)) +
  geom_freqpoly(binwidth = 1,
                aes(color = day)) +
  scale_x_continuous(breaks = 0:23) +
  scale_color_brewer(type = 'qual', palette = 2) +
  ggtitle('January 13th Pickups') +
  ylab('Number of Pickups')
# - Tues, Jan 13 (Typical Tuesday)
ggplot(subset(uber_jan, date == 13),
       aes(x = hour)) +
  geom_freqpoly(binwidth = 1,
                aes(color = day)) +
  scale_x_continuous(breaks = 0:23) +
  scale_color_brewer(type = 'qual', palette = 2) +
  ggtitle('January 13th Pickups') +
  ylab('Number of Pickups')
ggsave('jan_13_pickups.png')

# - Fri, Jan 30 (Busy Friday)
ggplot(subset(uber_jan, date == 31),
       aes(x = hour)) +
  geom_freqpoly(binwidth = 1,
                aes(color = day)) +
  scale_x_continuous(breaks = 0:23) +
  scale_color_brewer(type = 'qual', palette = 2) +
  ggtitle('January 31st Pickups') +
  ylab('Number of Pickups')
ggsave('jan_31_pickups.png')



# Testing corellation by hour and date
with(uber.by_hour, cor.test(hour, rides))
with(uber.by_date, cor.test(date, rides))

# Look at number of rides per day in one freq hist
ggplot(uber_jan,
       aes(x = date)) +
  geom_histogram(binwidth = 1,
                aes(fill = day)) +
  scale_x_continuous(breaks = 1:31) +
  scale_fill_brewer(type = 'seq', palette = 1) +
  ggtitle('January Pickups by Date') +
  ylab('Number of Pickups')
ggsave('january_pickups_by_date.png')

# Heatmap by month and date



# Best so far
ggplot(aes(y = date, 
           x = month),
       data = uber.by_date) +
  geom_tile(aes(fill = rides)) +
  scale_fill_gradientn(colours = colorRampPalette(c('blue', 
                                   'white', 
                                   'red'))(100),
                       guide = guide_colorbar(barheight = 20,
                                              title = 'Pickups')) +
  ggtitle('Rides per Date Heatmap') +
  xlab('Month') +
  ylab('Date') +
ggsave('rides_per_date_heatmap.png')

#Trying out scale_fill_gradientn
ggplot(aes(y = date, 
           x = month),
       data = uber.by_date) +
  geom_tile(aes(fill = rides)) +
  scale_fill_gradientn(colors = colorRampPalette(c(                                                   'blue',
                                  'aquamarine4',
                                  "gold",
                                  'firebrick'))(100),
                       guide = guide_colorbar(barheight = 20,
                                              title = 'Pickups')) +
  ylab('Date') +
  xlab('Month') +
  ggtitle('Rides per Date Heatmap')

# Keep going..
ggplot(aes(y = date, 
           x = month),
       data = uber.by_date) +
  geom_tile(aes(fill = rides)) +
  scale_fill_gradientn(colors = colorRampPalette(brewer.pal(11, 'RdYlBu'))(100),
                       guide = guide_colorbar(barheight = 20,
                                              title = 'Pickups')) +
  ylab('Date') +
  xlab('Month') +
  ggtitle('Rides per Date Heatmap')

ggplot(aes(y = date, 
           x = month),
       data = uber.by_date) +
  geom_tile(aes(fill = rides)) +
  scale_fill_gradientn(colors = colorRampPalette(brewer.pal(11, 'Spectral'))(100),
                       guide = guide_colorbar(barheight = 20,
                                              title = 'Pickups')) +
  ylab('Date') +
  xlab('Month') +
  ggtitle('Rides per Date Heatmap')

ggplot(aes(y = date, 
           x = month),
       data = uber.by_date) +
  geom_tile(aes(fill = rides)) +
  scale_colour_manual(values = rev(brewer.pal(11,"Spectral")),
guide = guide_colorbar(barheight = 20,
                                              title = 'Pickups')) +
  ylab('Date') +
  xlab('Month') +
  ggtitle('Rides per Date Heatmap')
```


# Density Map with stat_density2d

```{r}
nyc <- get_map("jackson heights, new york", zoom = 11)

nycMap <- ggmap(nyc,
                    extent = "device", 
                    legend = "none")

nycMap +
  stat_density2d(aes(x = lon, 
                     y = lat, 
                     fill = ..level..),
                 alpha = .2,
                 size = .5, 
                 bins = 50, 
                 data = subset(uber_sample, 
                               Borough != 'Unknown'),
                 geom = "polygon",
                 show.legend = F) +
  scale_fill_gradient2(#"Pickups",
                       #limits = c(0, 200),
                       low = "white", 
                       mid = "yellow", 
                       high = "red",
                       guide = 'none') +
  geom_point(aes(x = lon,
                 y = lat,
                 alpha = .05,
                 color = Borough),
             data = subset(uber_sample, 
                               Borough != 'Unknown'))

```

# Using binned location data

```{r}

nyc <- get_map("jackson heights, new york", zoom = 11)
nycMap <- ggmap(nyc,
                    extent = "device")

nycMap +
  geom_point(data = uber.by_loc,
             aes(x = lon,
                 y = lat,
                 color = n,
                 size = n/100,
                 alpha = .1)) +
  coord_trans

```



# Experiments Sections:
- Consider turning map scatterplot into map hexplot

```{r}
nycMap +
  coord_cartesian() +
  stat_binhex(data = subset(uber_sample,
                        Borough != 'Unknown'),
                 aes(x = lon,
                     y = lat,
                     alpha = .05))
?stat_density2d
```


```{r}
# NYC Map Scatter plot with density plot (no longer used)
nycMap +
  stat_density2d(aes(x = lon, 
                     y = lat, 
                     fill = ..level..),
                 alpha = .2,
                 bins = 10, 
                 data = subset(uber_sample, 
                               Borough != 'Unknown'),
                 geom = "polygon",
                 show.legend = F) +
  scale_fill_gradient2(#"Pickups",
                       low = "white", 
                       mid = "yellow", 
                       high = "red",
                       limits = c(0, 360),
                       breaks = seq(0, 350, 50)) +
  geom_point(aes(x = lon,
                 y = lat,
                 size = n,
                 color = n,
                 alpha = .01),
             #alpha = .01,
             data = subset(uber_sample.by_loc, 
                               Borough != 'Unknown')) +
  scale_color_gradient2(low = 'yellow',
                       high = 'red',
                       mid = 'yellow',
                       guide = guide_colorbar(barheight = 20,
                                              title = 'Pickups')) +
  scale_size(guide = 'none') +
  guides(alpha = F)

```

# (No longer nec) Transform data to get proper densities

This will limit your plots, so planning important. Graph each of the following
1. Bin the data by location
2. Bin the data by location + time:
  - Day
  - Hour
  - Month

```{r}

uber_sample.by_loc <- uber_sample %>%
  group_by(Zone) %>%
  summarise(n = n()) %>%
  inner_join(taxi_copy) %>%
  print

uber_jan.by_loc <- uber_jan %>%
  group_by(Zone) %>%
  summarise(rides = n()) %>%
  inner_join(taxi_copy) %>%
  print

uber_bigger_sample.by_loc <- uber_bigger_sample %>%
  group_by(Zone) %>%
  summarise(n = n()) %>%
  inner_join(taxi_copy) %>%
  print



```


# Sources of data:

1. Uber: https://github.com/fivethirtyeight/uber-tlc-foil-response
2. ggmap: D. Kahle and H. Wickham. ggmap: Spatial Visualization with ggplot2. 
The R Journal, 5(1), 144-161. URL
  http://journal.r-project.org/archive/2013-1/kahle-wickham.pdf

```{r}
save(taxi_lookup, 
     uber, 
     file = file.path('/Users/courtneyfergusonlee/p4/uber.RData'))
```

